Aldebaran~b's temperate past uncovered in planet search data

### 2018-02-05 paper/aldebaran.tex
gaussian process -> GP

\begin{abstract}
The nearby red giant Aldebaran is known to host a gas giant planetary companion from decades of ground-based spectroscopic measurements of its Doppler shift. Using Gaussian Process-based Continuous Auto-Regressive Moving Average (CARMA) models, we show that these historic data also contain evidence of acoustic oscillations in the star itself, and verify this result with further dedicated ground-based spectroscopy and space-based photometry with the \kepler Space Telescope. From the frequency of these oscillations we determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun, and note that this implies its planet will have been subject to insolation comparable to the Earth for some of the star's main sequence lifetime. Our approach to sparse, irregularly sampled time series astronomical observations has the potential to unlock asteroseismic measurements for thousands of stars in archival data, and push to lower-mass planets around red giant stars.
\end{abstract}

@@

Aldebaran ($\alpha$~Tauri) is a well-known first-magnitude naked-eye red giant star, and has long been the subject of astronomical investigations. It was one of the first stars around which an extrasolar planet candidate was identified, by looking for Doppler shifts from the star's reflex motion around the common centre of mass with its companion \citep[the radial velocity or RV method;][]{struverv}. While the hot Jupiter 51~Peg~b \citep{51peg} was the first exoplanet to be recognized as such, before this, \citet{hatzes1993} had noted RV variations in Pollux \citep[$\beta$~Gem; subsequently confirmed as a planet:][]{betgemconf,betgemconf2}, Arcturus ($\alpha$~Boo, unconfirmed), and Aldebaran. After further investigation by \citet{Hatzes1998}, \citet{Hatzes2015} have now claimed a firm RV detection of a planetary-mass companion Aldebaran~b, with a period of $629.96 \pm 0.90$~d.

@@

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process (GP), with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. Gaussian processes with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, these methods are impractical for long time series, as the computational
cost of evaluating the standard Gaussian Process likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the
solution of a class of stochastic ordinary differential equations, the
Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can
be evaluated in linear time. For problems such as this, we therefore have access
to these powerful computational tools for inferring their power spectral
densities.  Our treatment of CARMA models is described more fully in Appendix
\ref{carmatheory}.   \citet{Kelly2014} also give more details on CARMA
processes in an astronomical context, and show how a state-space model of the
process $y(t)$ can be tracked through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of
zero mean and known (heteroskedastic) variance, using a Kalman filter to produce
a likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on
the amplitude of the stochastic forcing, $\sigma$, the (possibly complex)
eigenfrequencies of the ODE, $r_i$, and parameters describing the power spectrum
of the stochastic forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational
time.  Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard Gaussian Process covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

@@

A CARMA process corresponds to the solution, $y(t)$, of the stochastic ODE
%
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
  \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
%
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process;
$r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation
of the ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the
inverse process; and $\eta(t)$ is a white-noise Gaussian Process with
$\left\langle \eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t')
\right\rangle = \sigma^2 \delta\left(t - t' \right)$.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is
%
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
%
The power spectrum is a rational function of frequency with poles at $2\pi i f =
r_i$ and zeros at $2\pi i f = b_i$; because of the invertable constraints, these
poles and zeros all occur when $\Im f \neq 0$.  The autocorrelation function
corresponding to the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
%
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
%
where the RMS amplitudes of the $p$ different modes, $A_i$, are functions of
$\sigma$ and the values of the roots $r_i$ and $b_j$.  If $q = p-1$, then the
$A_i$ are independent, while if $q < p-1$ then there are interdependencies among
the $A_i$.  From either the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the
autocorrelation function in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a
real root $r_i$ represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an oscillatory mode
with decay rate $\Re r_i$ and angular frequency $\left|\Im r_i\right|$.
Oscillatory modes can also be described by their frequency $f$ and quality
factor $Q$ via $r = f/(2Q) \pm 2\pi i f$.  Decaying modes generate a damped
random walk, also known as a Ornstein-Uhlenbeck process.  More discussion of CARMA processes, including formulae for $A_i$ in terms of $\sigma$, $r_i$, and $b_j$, can be found in \citet{Kelly2014} and references therein.

@@

As a final step to reduce residual uncorrected systematics, we apply the \textsc{k2sc} \citep{k2sc} Gaussian Process-based systematics correction code to the initial halo light curve, but the effect of this in the present case for Aldebaran is minimal.
There is somewhat higher than usual residual noise at harmonics of $46\,\mu\mathrm{Hz}$ ($4\,\mathrm{d}^{-1}$, the satellite thruster firing frequency), but this is nevertheless very small in comparison to the signal from Aldebaran, and may be ascribed to the large fraction of the pixel mask occupied by the bleed column from this extremely bright star.

\section{Stellar Modelling}\label{sec:mod}
\subsection{Stellar Models}\label{sec:stell_mod}
We used our determination of \numax and several combinations of the asteroseismic and spectroscopic parameters, along with luminosity, to estimate the fundamental stellar parameters, via fitting to stellar models. We used \textsc{MESA} models \citep{2011Paxton,2013Paxton} in conjunction with the Bayesian code \textsc{PARAM} \citep{2006dasilva, 2017Rod}.  A summary of our selected ``benchmark'' options is as follows;
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}%GUY THIS BIT COULD PROBABLY BE DROPPED AND REFERENCED OUT TO \citep{2017Rod}
\item Heavy element partitioning from \cite{1993Grevesse}.
\item OPAL equation of state \citep{2002Rogers} along with OPAL opacities \citep{1996Iglesias}, with complementary values at low temperatures from \cite{2005Ferguson}.
\item Nuclear reaction rates from NACRE \citep{1999Angulo}.
\item The atmosphere model according to \cite{1966Kris}.
\item The mixing length theory to describe convection (we adopt a solar-calibrated parameter $\alpha_{\textrm{MLT}} =1.9657$).
\item Convective overshooting on the main sequence is set to $\alpha_{\textrm{ov}}=0.2H_{p}$, with $H_{p}$ the pressure scale height at the border of the convective core. Overshooting was applied according to the \cite{1975Maeder} step function
scheme.
\item We consider only the red giant branch phase to avoid the treatment of poorly understood mass loss at the tip of the red giant branch.
\item No rotational mixing or diffusion is included.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We do not need to correct for the line-of-sight Doppler shift at the frequency precision available in our data \citep{2014MNRAS.445L..94D}.
\subsection{Additional modelling inputs}\label{sec:addmod}
In addition to the asteroseismic parameters, spectroscopically-determined temperature and metallicity values are needed. There exist multiple literature values for Aldebaran. We chose to compare a range of literature values to investigate what uncertainty these systematically-differing models produce in inferred stellar properties.

>>>>

\begin{abstract}
The nearby red giant Aldebaran is known to host a gas giant planetary companion from decades of ground-based spectroscopic radial velocity measurements. Using Gaussian Process-based Continuous Auto-Regressive Moving Average (CARMA) models, we show that these historic data also contain evidence of acoustic oscillations in the star itself, and verify this result with further dedicated ground-based spectroscopy and space-based photometry with the \kepler Space Telescope. From the frequency of these oscillations we determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun, and note that this implies its planet will have been subject to insolation comparable to the Earth for some of the star's main sequence lifetime. Our approach to sparse, irregularly sampled time series astronomical observations has the potential to unlock asteroseismic measurements for thousands of stars in archival data, and push to lower-mass planets around red giant stars.
\end{abstract}

@@

Aldebaran ($\alpha$~Tauri) is a well-known first-magnitude red giant star, and has long been the subject of astronomical investigations. It was one of the first stars around which an extrasolar planet candidate was identified, by looking for Doppler shifts from the star's reflex motion around the common centre of mass with its companion \citep[the radial velocity or RV method;][]{struverv}. While the hot Jupiter 51~Peg~b \citep{51peg} was the first exoplanet to be recognized as such, before this, \citet{hatzes1993} had noted RV variations in Pollux \citep[$\beta$~Gem; subsequently confirmed as a planet:][]{betgemconf,betgemconf2}, Arcturus ($\alpha$~Boo, unconfirmed), and Aldebaran. After further investigation by \citet{Hatzes1998}, \citet{Hatzes2015} have now claimed a firm RV detection of a planetary-mass companion Aldebaran~b, with a period of $629.96 \pm 0.90$~d.

@@

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process (GP), with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. GPs with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, these methods are impractical for long time series, as the computational
cost of evaluating the standard GP likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the
solution of a class of stochastic ordinary differential equations, the
Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can
be evaluated in linear time. For problems such as this, we therefore have access
to these powerful computational tools for inferring their power spectral
densities.  Our treatment of CARMA models is described more fully in Appendix
\ref{carmatheory}.   \citet{Kelly2014} also give more details on CARMA
processes in an astronomical context, and show how a state-space model of the
process $y(t)$ can be tracked through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of
zero mean and known (heteroskedastic) variance, using a Kalman filter to produce
a likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on
the amplitude of the stochastic forcing, $\sigma$, the (possibly complex)
eigenfrequencies of the ODE, $r_i$, and parameters describing the power spectrum
of the stochastic forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational
time.  Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard GP covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard GP likelihood function in
$\mathcal{O}\left( n \right)$ time.

@@

A CARMA process corresponds to the solution, $y(t)$, of the stochastic ODE
%
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
  \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
%
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process;
$r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation
of the ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the
inverse process; and $\eta(t)$ is a white-noise GP with
$\left\langle \eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t')
\right\rangle = \sigma^2 \delta\left(t - t' \right)$.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is
%
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
%
The power spectrum is a rational function of frequency with poles at $2\pi i f =
r_i$ and zeros at $2\pi i f = b_i$; because of the invertable constraints, these
poles and zeros all occur when $\Im f \neq 0$.  The autocorrelation function
corresponding to the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
%
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
%
where the RMS amplitudes of the $p$ different modes, $A_i$, are functions of
$\sigma$ and the values of the roots $r_i$ and $b_j$.  If $q = p-1$, then the
$A_i$ are independent, while if $q < p-1$ then there are interdependencies among
the $A_i$.  From either the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the
autocorrelation function in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a
real root $r_i$ represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an oscillatory mode
with decay rate $\Re r_i$ and angular frequency $\left|\Im r_i\right|$.
Oscillatory modes can also be described by their frequency $f$ and quality
factor $Q$ via $r = f/(2Q) \pm 2\pi i f$.  Decaying modes generate a damped
random walk, also known as a Ornstein-Uhlenbeck process.  More discussion of CARMA processes, including formulae for $A_i$ in terms of $\sigma$, $r_i$, and $b_j$, can be found in \citet{Kelly2014} and references therein.

@@

As a final step to reduce residual uncorrected systematics, we apply the \textsc{k2sc} \citep{k2sc} GP-based systematics correction code to the initial halo light curve, but the effect of this in the present case for Aldebaran is minimal.
There is somewhat higher than usual residual noise at harmonics of $46\,\mu\mathrm{Hz}$ ($4\,\mathrm{d}^{-1}$, the satellite thruster firing frequency), but this is nevertheless very small in comparison to the signal from Aldebaran, and may be ascribed to the large fraction of the pixel mask occupied by the bleed column from this extremely bright star.

\section{Stellar Modelling}\label{sec:mod}
\subsection{Stellar Models}\label{sec:stell_mod}
We used our determination of \numax and several combinations of the asteroseismic and spectroscopic parameters, along with luminosity, to estimate the fundamental stellar parameters, via fitting to stellar models. We used \textsc{MESA} models \citep{2011Paxton,2013Paxton} in conjunction with the Bayesian code \textsc{PARAM} \citep{2006dasilva, 2017Rod}.  A summary of our selected ``benchmark'' options is as follows;
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}%GUY THIS BIT COULD PROBABLY BE DROPPED AND REFERENCED OUT TO \citep{2017Rod}
\item Heavy element partitioning from \cite{1993Grevesse}.
\item OPAL equation of state \citep{2002Rogers} along with OPAL opacities \citep{1996Iglesias}, with complementary values at low temperatures from \cite{2005Ferguson}.
\item Nuclear reaction rates from NACRE \citep{1999Angulo}.
\item The atmosphere model according to \cite{1966Kris}.
\item The mixing length theory to describe convection (we adopt a solar-calibrated parameter $\alpha_{\textrm{MLT}} =1.9657$).
\item Convective overshooting on the main sequence is set to $\alpha_{\textrm{ov}}=0.2H_{p}$, with $H_{p}$ the pressure scale height at the border of the convective core. Overshooting was applied according to the \cite{1975Maeder} step function
scheme.
\item No rotational mixing or diffusion is included.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We do not need to correct for the line-of-sight Doppler shift at the frequency precision available in our data \citep{2014MNRAS.445L..94D}.
\subsection{Additional modelling inputs}\label{sec:addmod}
In addition to the asteroseismic parameters, spectroscopically-determined temperature and metallicity values are needed. There exist multiple literature values for Aldebaran. We chose to compare a range of literature values to investigate what uncertainty these systematically-differing models produce in inferred stellar properties.
